On the spatial aggregation of ecosystem condition indicators

### 2025-01-29 manuscript/manuscript.qmd
sea ch

The term spatial aggregation is also sometimes used to refer to **upscaling** in the sense of taking local information and making it applicable or translate it to larger, e.g. landscape or regional, scales [@knotters2010; @finke2002].
This process often involves predicting values across larger areas or surfaces, generally through models, including krieging-based models and distribution models.
When the data (e.g. indicator values) are made wall-to-wall, one can then more easily get point estimates for larger areas via simple aggregation functions such as summation or by taking the mean.
Using models to predict variable or indicator values is allowed under the SEEA EA. However, in this paper we are addressing the second part of this process only, of simplifying the spatial variation, from many to few data points.

@@

SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2021 @unitednations2022].
We will focus on the SEEA EA condition accounts in this paper.
In these accounts, information about ecosystem condition is made spatially explicit by assigning **variable** or **indicator** values to individual ecosystem assets.
Unless the underlying data are wall-to-wall maps, for example from satellite images, then this process of assigning values to ecosystem assets already here requires some spatial manipulations in the form of aggregation, dissaggregation, extrapolation or interpolation (ref Ewert 2011?), of which averaging or summing are the most common methods.
Subsequently, information from ecosystem assets may be spatially aggregated to larger areas, such as the entire ecosystem accounting area.
When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels.
The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.

In ECAs, data are often not wall-to-wall, and when upscaling is not done, as defined above, the spatial aggregation needs to account for the spatial structure of the data in a different way, such as design-based approached (relying on random or balanced samples across the entire EEA), or a step-wise aggregation via ecosystem assets, homogenous ecosystem areas [@vallecillo2022] or homogenous impact areas [@kolstad_making_2024].
Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means.
When using means, area weighting is required when following SEEA EA.

@@

# Compliance across nested assessments

>>>>

The term spatial aggregation is also sometimes used to refer to **upscaling** in the sense of taking local information and making it applicable or translate it to larger, e.g. landscape or regional, scales [@knotters2010; @finke2002].
This process often involves predicting values across larger areas or surfaces, generally through models, including krieging-based models and distribution models.
When the data (e.g. indicator values) are made wall-to-wall, one can then more easily get point estimates for larger areas via simple aggregation functions such as summation or by taking the mean.
Using models to predict variable or indicator values is allowed under the SEEA EA. However, in this paper we are addressing the second part of this process only, of simplifying the spatial variation, from many to few data points (sometimes also included in the terms upscaling).

@@

SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2021 @unitednations2022].
We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account.

In ECAs, information about ecosystem condition is made spatially explicit by assigning **variable** or **indicator** values to individual ecosystem assets.
The process of assigning values to ecosystem assets usually requires some spatial manipulations of data, or scale changes [@ewert2011].
The type of manipulation depends on the spatial resolution of the underlying data, as well as the size and configuration of the ecosystem assets.
To main types of data can be distinguished: sparse data and complete data.

Complete, or wall-to-wall data, typically include remote sensing products.
It can also include other types of pre-aggregated data, where sparse data, such as field survey summaries, are assigned to areal units.
This is very common.
A lot of demographic information for example is made available at the scale of administrative units, rather than at the individual level (which would be very ineffective and problematic also in terms of protecting sensitive information).
Similar pre-aggregation is, however, ambiguous and true for most complete data.
Consider as an example data from LiDAR (Light Detection And Ranging).
The raw data is a point cloud, yet the most commonly available datasets are surface or canopy models; wall-to-wall generalisations of the point cloud.
Also data such as rasters derived from satellite images, such as NDVI (extensively used in ECAs), have some inherent pre-aggregation where values are averaged for each raster cell.
Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (@nte-rel).
This issue will affect our ability to detect true maximum or minimum values (as used for *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step.
This will clearly affect the suitability of such data in looking for areas of high risk (very degraded areas, for example).
The problem is the same when we deliberately aggregate data later in the indicator development process, and we always need to ensure that the spatial aggregation function we use does not have any unintentional effect on the type of inference that we can draw from the metric.

For sparse data we are considering type point data, with variable values assigned to singular points or small areas.
In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
One can assign values to areas via an aggregation functions used on data values that fall within a given ecosystem asset.
Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means.
When using means, area weighting is required when following SEEA EA. This is what we refer to as deliberate spatial aggregation, and what is the main focus of this paper.
When doing this, we need to account for the spatial structure (spatial correlations) of the data (@sec-auto).
One way is to rely on design-based approaches (relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets.
In our experience, the choice of which areal units to use when doing this step-wise aggregating is often not reflected much upon.
It may be a limitation with the data related to the type of pre-aggregation performed (see above).
In some cases the areal units are decided politically, wit little scientific reasoning behind it.
However, the arbitrariness of this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
It arguable important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods (REF).
The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].

Interpolation or other modeling can also be used to impute or predict data values between the points where we have information.
However, this is not the focus of this paper, and in our experience this has not been common practice in SEEA EA accounts produced to date.

When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy, as well as many more concrete pitfalls in spatial aggregation of information.
The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.

@@

# Compliance across nested assessments {#sec-comp}
